Open source symbolic multibody dynamics software

### 2012-02-07 PetersonGedeHubbard2012.tex
Added a bit about permutation matrices.

Equating the right hand side of equations (\ref{eq:f0_plus_f1_expansion}) and
(\ref{eq:f2_plus_f3_expansion}) each to zero, eliminating $\bm{\dot{q}}$ from
(\ref{eq:f2_plus_f3_expansion}), and rewriting in matrix form, we obtain
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      M_{00} & \bm{0}_{n \times o} \\
      \bm{0}_{o \times n} & M_{11}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       A_{00} & A_{01} \\
       A_{10} & A_{11}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{n \times q} \\
        B_{1}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet as was discussed earlier, only $p = n - l + o - m$ of these
quantities are independent.  To address this issue, one must pick a particular
set of independent coordinates and speeds and appeal to equations
(\ref{eq:holonomic_expansion}) and (\ref{eq:nonholonomic_expansion}).  To this
end, consider the following partitioning of the generalized coordinates and
generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.

Setting the right hand side of equation (\ref{eq:holonomic_expansion}) equal to
zero, and making use of $P_q$, we have
\begin{align}
  \nabla_{\bm{q}}\bm{f}_{h} P_{q} \delta \bm{\tilde{q}} &=
  \nabla_{\bm{q}}\bm{f}_{h} P_{qi} \delta \bm{q_i} +
  \nabla_{\bm{q}}\bm{f}_{h} P_{qd} \delta \bm{q_d}\notag\\
  &= \bm{0}\notag\\
  \implies \delta \bm{q}_d &= -(\nabla_{\bm{q}}\bm{f}_{h} P_{qd})^{-1}
  (\nabla_{\bm{q}}\bm{f}_{h} P_{qi}) \delta \bm{q}_i
\end{align}

>>>>

Equating the right hand side of equations (\ref{eq:f0_plus_f1_expansion}) and
(\ref{eq:f2_plus_f3_expansion}) each to zero, eliminating $\bm{\dot{q}}$ from
(\ref{eq:f2_plus_f3_expansion}), and rewriting in matrix form, we obtain
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      M_{00} & \bm{0}_{n \times o} \\
      \bm{0}_{o \times n} & M_{11}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       A_{00} & A_{01} \\
       A_{10} & A_{11}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{n \times q} \\
        B_{1}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet  only $p = n - l + o - m$ of these quantities are independent, due to
the constraints (equations (\ref{eq:holonomic} and (\ref{eq:nonholonomic})).
To address this issue, one must pick a particular set of independent
coordinates and speeds and appeal to equations (\ref{eq:holonomic_expansion})
and (\ref{eq:nonholonomic_expansion}).  To this end, consider the following
partitioning of the generalized coordinates and generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.  We use the notation $P_{qi}$
and $P_{qd}$ to denote the first $n-l$ and last $l$ columns of $P_q$,
respectively; similarly, $P_{ui}$ is the first $o-m$ colums of $P_{u}$ while
$P_{ud}$ is the last $m$ columns of $P_u$.

Setting the right hand side of equation (\ref{eq:holonomic_expansion}) equal to
zero, and making use of $P_q$, we have
\begin{align}
  \nabla_{\bm{q}}\bm{f}_{h} P_{q} \cdot \delta \bm{\tilde{q}} &=
  \nabla_{\bm{q}}\bm{f}_{h} P_{qi} \cdot \delta \bm{q_i} +
  \nabla_{\bm{q}}\bm{f}_{h} P_{qd} \cdot \delta \bm{q_d}\notag\\
  &= \bm{0}\notag\\
  \implies \delta \bm{q}_d &= -(\nabla_{\bm{q}}\bm{f}_{h} P_{qd})^{-1}
  (\nabla_{\bm{q}}\bm{f}_{h} P_{qi}) \cdot \delta \bm{q}_i \notag\\
  \implies \delta \bm{q} &= \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{h} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{h} \right] P_{qi} \cdot \delta
    \bm{q}_i
\end{align}
For convenience, we define
\begin{equation}
  \label{eq:C_0}
  C_0 \triangleq \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{h} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{h} \right] P_{qi}
\end{equation}
which enables us to write
\begin{align}
  \label{eq:delta_q}
  \delta \bm{q} &= C_0 \cdot \delta \bm{q}
\end{align}

Applying the same approach to (\ref{eq:nonholonomic_expansion}), we obtain
\begin{align}
\nabla_{\bm{q}}f_{nh} \cdot \delta \bm{q} +
\nabla_{\bm{u}}f_{nh} \cdot \delta \bm{u} &= \nabla_{\bm{q}} \bm{f}_{nh} \cdot
\delta \bm{q} + \nabla_{\bm{u}} \bm{f}_{nh} P_{ui} \cdot \delta \bm{u}_i +
\nabla_{\bm{u}} \bm{f}_{nh} P_{ud} \cdot \delta \bm{u}_d \notag\\
&= \bm{0}\notag\\
\implies \delta \bm{u}_d &= -\left(\nabla_{\bm{u}} \bm{f}_{nh}
P_{ud}\right)^{-1}\left[\nabla_{\bm{q}}\bm{f}_{nh} \cdot \delta\bm{q} +
  \nabla_{\bm{u}} \bm{f}_{nh} P_{ui} \cdot \delta \bm{u}_i \right]\notag
\end{align}
We define
\begin{align}
  \label{eq:C_1}
  C_1 &\triangleq -\left(\nabla_{\bm{u}} \bm{f}_{nh} P_{ud}\right)^{-1}
  \nabla_{\bm{q}}\bm{f}_{nh} \\
  \label{eq:C_2}
  C_2 &\triangleq -\left(\nabla_{\bm{u}} \bm{f}_{nh} P_{ud}\right)^{-1} \nabla_{\bm{u}} \bm{f}_{nh} P_{ui}
\end{align}
which enables us to write
\begin{align}
  \label{eq:delta_u}
  \delta \bm{u} &= C_1 \cdot \delta \bm{q} + C_2 \cdot \delta \bm{u}_i\notag\\
  &= C_1 C_0 \cdot \delta \bm{q}_i + C_2 \cdot \delta \bm{u}_i
\end{align}

By making use of equations (\ref{eq:delta_q}) and (\ref{eq:delta_u}), we can
rewrite equation (\ref{eq:state_space_unconstrained}) as
\begin{align}
  \label{eq:state_space_constrained}
  \left[
    \begin{array}{cc}
      M_{00} & \bm{0}_{n \times o} \\
      \bm{0}_{o \times n} & M_{11}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       (A_{00} + A_{01} C_{1}) C_{0} & A_{01} C_{2} \\
       (A_{10} + A_{11} C_{1}) C_{0} & A_{11} C_{2}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q}_i \\
        \delta \bm{u}_i
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{n \times q} \\
        B_{1}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}

There are two requirements for this method to work, namely, both
$\nabla_{\bm{q}}\bm{f}_{h} P_{qd}$ and $\nabla_{\bm{u}} \bm{f}_{nh} P_{ud}$
must be nonsingular.  Because of the presence of the permutation matrices, this
implies proper selection of dependent coordinates and dependent speeds is
critical.  The permutation matrices effectively select the columns of the
constraint gradient matrices, and may be selected by examination of the
constraint gradient matrix through a SVD or other numerical techniques; for
small systems it may be possible to choose dependent speeds without resorting
to numerical techniques if the system trajectories stay with the configuration
space where these matrices are nonsingular.

For stability analyses, one must examine the stability of the mapping $\delta
\bm{q}_i \to \bm{\dot{q}}_i$.  This can be done by first solving for
$\delta\bm{\dot{x}} = \left[\delta\bm{\dot{q}},\,\delta\bm{\dot{u}}\right]^{T}$, then selecting
the independent rows of these equations ($n-l+o-m$ of them) to form a square
matrix whose eigenvalues govern the stability of the system about the point of
linearization.
